Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-16 manuscript/plos/review_1.tex
shorter abstract

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses. They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network. Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility. We applied them to GENESIS, a GWAS on French women with a family history of breast cancer and who tested negative for pathogenic variants in \emph{BRCA1} and \emph{BRCA2}. We critically compared these six methods, discussing the impact of their different mathematical frameworks, and the parameter choices. Additionally, we performed an in-depth benchmarking with respect to the size and predictive power of their solutions, as well as their stability and runtimes. Importantly, we found significant overlaps between the genes in five of the solution networks and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. Since most of the methods produced reasonable solutions, we proposed to combine them into a consensus solution, containing the genes selected by at least two methods. This aggregation brought further insights. For instance, it contained \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods with regards to conventional $\chi^2$-based GWAS was instability i.e. network methods' outputs changed more in front of small perturbations in the input. To compensate for this instability, we proposed a stable consensus solution, formed by the most consistently selected genes in different subsamples of the data. This stable consensus was composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, Fisher's exact test P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network was organized around \emph{CUL3}, which encodes a ubiquitin ligase related protein that regulates the protein levels of several genes involved in cancer progression. In conclusion, we showed how network address limitations of GWAS, namely their lack of statistical power and the difficulty of their interpretation. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
% Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses.
% They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network.
Network approaches to disease use molecular networks, which model functional relationships between the molecules in a cell, to generate hypotheses about the genetics of complex diseases.
Among them, several approaches jointly consider gene scores, representing the association between each gene and the disease, and the biological context of each gene, modeled by a network.
%These approaches can select low-scoring genes if they connect multiple high-scoring genes.
Here, we obtain gene scores from genome-wide association studies (GWAS) and study six network approaches on 
% Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility.
% We apply them to
GENESIS, a GWAS on French women with non-BRCA familial breast cancer% and no pathogenic variants in \emph{BRCA1} and \emph{BRCA2}
. We provide a critical comparison of these six methods, discussing the impact of their mathematical formulation and of their parameters. % on % . Additionally, we performed an in-depth benchmarking with respect to
% the size and predictive power of their solutions, as well as on their stability and runtimes.
% Importantly,
Using a molecular network yields more compelling results than standard % SNP- and gene-level
GWAS analyses. Indeed, %recovering causal subnetworks tightly related to cancer susceptibility.
we find significant overlaps between our solutions and the genes identified in the largest GWAS on breast cancer susceptibility. % Since most of the methods produce reasonable solutions, we
We further
propose to combine these solutions into a consensus network, %containing the genes selected by at least two methods. This aggregation
which
brings further insights. The consensus network contains \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods % with respect to conventional $\chi^2$-based GWAS
is that they are not robust to small perturbations in their inputs. %To compensate for this instability, we
We therefore propose a stable consensus solution, formed by the most consistently selected genes in multiple subsamples of the data. In GENESIS, it is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network is organized around \emph{CUL3}, which is involved in the regulation of several genes linked to cancer progression. In conclusion, we showed how network methods help overcome the lack of statistical power of GWAS and improve their interpretation. Project-agnostic implementations of all methods are available at \url{https://github.com/hclimente/gwas-tools}. % to facilitate their application to other GWAS datasets.
